miniKanren with fair search strategies

### 2019-05-06 Paper/paper.tex
sec DFSi

@@

miniKanren is a family of relational programming languages.
\citet{friedman_reasoned_2018} introduce miniKanren and its implementation
in the The Reasoned Schemer, 2nd Ed(TR2). miniKanren programs, especially
relational interpreters, have been proven to be useful in solving many problems 
by \citet{byrd2017unified}.

% \conde is the flagship disjunction operator in miniKanren.
A subtlety arises when a \conde contains many clauses: not every clause has the 
equal chance to contribute to the result. As an example, consider the following 
relation \texttt{repeato} and its invocation.

@@

Next, consider the following disjuction of invoking \texttt{repeato} with four 
different letters.

@@

\conde intuitively relates its clauses with logical \texttt{or}. And thus an 
unsuspicious beginner would expect each letter to contribute equally to the 
result, as following.

@@

The miniKanren in TR2 implements interleaving DFS(\DFSi), the cause of this 
unexpected result. With this search strategy, each clause takes a half 
of its received computational resources and pass the other half to its 
following clauses, except for the last clause that takes all resources it 
receives. In the example above, the \texttt{a} clause takes half of all 
recourses. And the \texttt{b} clause takes a quarter. Thus \texttt{c} and 
\texttt{d} barely contribute to the result.

@@

There is another reason that miniKanren could use more search strategies than
just \DFSi. In many applications, there does not exist one order that serves
for all purposes. For example, a relational dependent type checker contains
clauses for constructors that build data and clauses for eliminators that use
data. When this type checker is used to generate simple and shallow programs,
the clauses of constructors should be put in the front of \conde.
When performing proof search for complicated programs, the clauses of eliminator
should take the focus. With \DFSi, these two uses can not be efficient at the
same time. In fact, to make one use efficient, the other one must be drastically
slow.

The specification that every clause in the same \conde{} is given equal 
``search priority'' is called fair \disj{}. And search strategies with 
almost-fair \disj{} give every clauses in the same \conde{} similar priority. 
Fair \conj{}, a related concept, is more complicated. We defer it to the next 
section.

To summarize our contribution, we
\begin{itemize}
	\item propose and implement balanced interleaving depth-first search 
	(DFS$_\textrm{bi}$), a new search strategy with almost-fair \disj{}.
	\item propose and implement fair depth-first search (DFS$_\textrm{f}$), a 
	new search strategy with fair \disj{}.
	\item implement in a new way breath-first search (BFS), a search strategy 
	with fair \disj{} and fair \conj{}. (our code runs faster in all 
benchmarks and is simpler). And we prove formally that our BFS implementation 
is equivalent to the one by \citet{seres1999algebra}.
\end{itemize}

@@

In this section, we define fair \disj{}, almost-fair \disj{} and fair \conj{}. 
Before going further into fairness, we would like to give a short review about 
state, search space, and goal, because fairness is defined in terms of them. 
A \emph{state} is a collection of constraints. Every answer given by 
miniKanren corresponds to a state. A \emph{search space} is a collection of 
states. And a \emph{goal} is a function from a state to a search space.

@@

\begin{defn}[fair \disj{}]
A \disj{} is fair iff it allocates computational resource evenly to 
search spaces produced by goals in the same disjunction (i.e. clauses in 
the same \conde).
\end{defn}

\begin{defn}[almost-fair \disj{}]
A \disj{} is almost-fair iff it allocates computational resource so evenly to 
search spaces produced by goals in the same disjunction that the maximal ratio 
of resources is bounded by a constant.
\end{defn}

\begin{defn}[unfair \disj{}]
A \disj{} is unfair iff it is not even almost-fair.
\end{defn}

@@

@@

\begin{center}
	\begin{tabular}{c|c}
		\lstinputlisting{Figures/repeato-conj-fair.rkt}
	\end{tabular}
\end{center}

@@

In this section we review the implementation of interleaving depth-first search 
(\DFSi). ... ...

@@

Balanced interleaving DFS (DFS$_\textrm{bi}$) has almost-fair \disj{} and unfair \conj{}. 
The implementation of DFS$_\textrm{bi}$ differs from DFS$_\textrm{i}$ in the \disj{} macro. We list the 
new \disj{} with its helpers in Fig.~\ref{balanced-disj}. The first helper 
function, \texttt{split}, takes a list of goals \texttt{ls} and a procedure 
\texttt{k}, partitions \texttt{ls} into two sub-lists of roughly equal length, 
and returns the application of \texttt{k} to the two sub-lists. \texttt{disj*} 
takes a non-empty list of goals \texttt{gs} and returns a goal. With the help 
of \texttt{split}, it essentially constructs a \emph{balanced} binary tree 
where leaves are elements of \texttt{gs} and nodes are \texttt{disj2}s, hence 
the name of this search strategy. In contrast, the \disj{} in DFS$_\textrm{i}$
constructs the binary tree with the same nodes but in an unbalanced form.

@@

Fair DFS (DFS$_\textrm{f}$) has fair \disj{} and unfair \conj{}. The implementation of DFS$_\textrm{f}$ 
differs from DFS$_\textrm{i}$'s in \texttt{disj2} (Fig.~\ref{fDFS}). \texttt{disj2} is 
changed to call a new and fair version of \texttt{append-inf}. 
\texttt{append-inf/fair} immediately calls its helper, \texttt{loop}, with the 
first argument, \texttt{s?}, set to \texttt{\#{}t}, which indicates that 
\texttt{s-inf} and \texttt{t-inf} haven't been swapped. The swapping happens at 
the third \texttt{cond} clause in the helper, where \texttt{s?} is updated 
accordingly. The first two \texttt{cond} clauses essentially copy the 
\texttt{car}s and stop recursion when one of the input spaces is obviously 
finite. The third clause, as we mentioned above, is just for swapping. When the 
fourth and last clause runs, we know that both \texttt{s-inf} and \texttt{t-inf} 
are ended with a thunk. In this case, a new thunk is constructed. The new thunk 
calls the driver recursively. Here changing the order of \texttt{t-inf} and 
\texttt{s-inf} won't hurt the fairness (though it will change the order of 
answers). We swapped them back so that answers are produced in a more natural 
order.

@@

BFS has both fair \disj{} and fair \conj{}. Our implementation is based on 
DFS$_\textrm{f}$ (not DFS$_\textrm{i}$). To implement \BFS{} based on \DFSf{} 
all we have to do is renaming 
\texttt{append-map-inf} to \texttt{append-map-inf/fair}, then replacing its use 
of \texttt{append-inf} to \texttt{append-inf/fair}.

>>>>

\NOTE[LKC]{\disj{} and \conj{} occurs free in the abstract.}

@@

miniKanren is a family of relational programming languages.
\citet{friedman_reasoned_2018} introduce miniKanren and its implementation
in \emph{The Reasoned Schemer, 2nd Ed} (TR2). miniKanren programs, especially
relational interpreters, have been proven to be useful in solving many problems 
by \citet{byrd2017unified}.

% \conde is the flagship disjunction operator in miniKanren.
A subtlety arises when a \conde contains many clauses: not every clause has an 
equal chance to contribute to the result. As an example, consider the following 
relation \texttt{repeato} and its invocation.

@@

Next, consider the following disjunction of invoking \texttt{repeato} with four 
different letters.

@@

\conde intuitively relates its clauses with logical \texttt{or}. And thus an 
unsuspicious beginner would expect each letter to contribute equally to the 
result, as follows.

@@

The miniKanren in TR2 implements interleaving DFS(\DFSi), the cause of this 
unexpected result. With this search strategy, each clause takes half 
of its received computational resources and pass the other half to its 
following clauses, except for the last clause that takes all resources it 
receives. In the example above, the \texttt{a} clause takes half of all 
recourses. And the \texttt{b} clause takes a quarter. Thus \texttt{c} and 
\texttt{d} barely contribute to the result.

@@

There is another reason that miniKanren could use more search strategies than
just \DFSi. In many applications, there does not exist one order that serves
for all purposes. For example, a relational dependent type checker contains
clauses for constructors that build data and clauses for eliminators that use
data. When the type checker is used to generate simple and shallow programs,
the clauses of constructors should be put in the front of \conde.
When performing proof searches for complicated programs, the clauses of 
eliminator should take the focus. With \DFSi, these two uses cannot be 
efficient at the same time. In fact, to make one use efficient, the other one 
must be drastically slow.

The specification that every clause in the same \conde{} is given equal 
``search priority'' is called fair \disj{}. And search strategies with 
almost-fair \disj{} give every clause in the same \conde{} similar priority. 
Fair \conj{}, a related concept, is more complicated. We defer it to the next 
section.

To summarize our contribution, we
\begin{itemize}
	\item propose and implement balanced interleaving depth-first search 
	(\DFSbi{}), a new search strategy with almost-fair \disj{}.
	\item propose and implement fair depth-first search (\DFSf{}), a 
	new search strategy with fair \disj{}.
	\item implement in a new way breath-first search (BFS), a search strategy 
	with fair \disj{} and fair \conj{}. (our code runs faster in all 
benchmarks and is simpler). And we prove formally that our BFS implementation 
is equivalent to the one by \citet{seres1999algebra}.
\end{itemize}

@@

In this section, we define fair \disj{}, almost-fair \disj{} and fair \conj{}. 
Before going further into fairness, we would like to give a short review about 
state, search space, and goal, because fairness is defined in terms of them. 
A \emph{state} is a collection of constraints. Every answer corresponds to a 
state. A \emph{search space} is a collection of states. And a \emph{goal} is a 
function from a state to a search space.

@@

\begin{defn}[fair \disj{}]
A \disj{} is fair if and only if it allocates computational resource evenly to 
search spaces produced by goals in the same disjunction (i.e. clauses in 
the same \conde).
\end{defn}

\begin{defn}[almost-fair \disj{}]
A \disj{} is almost-fair if and only if it allocates computational resource so 
evenly to 
search spaces produced by goals in the same disjunction that the maximal ratio 
of resources is bounded by a constant.
\end{defn}

\begin{defn}[unfair \disj{}]
A \disj{} is unfair if and only if it is not even almost-fair.
\end{defn}

@@

\NOTE[LKC]{Checked grammer of text before this line}

@@

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-conj-fair.rkt}
	\end{tabular}
\end{center}

@@

In this section we review the implementation of interleaving depth-first search 
(\DFSi). This review is for comparison with other search strategies in this 
paper. Thus we focus on parts that are changed later. TRS2 
\citep{friedman_reasoned_2018} provides a comprehensive description of the 
whole miniKanren implementation.

\begin{figure}
    \lstinputlisting{Figures/DFSi.rkt}
    \caption{implementation of \DFSi}
    \label{DFSi}
\end{figure}

Fig.~\ref{DFSi} shows part of the implementation of \DFSi{}. We follow a
convention to name variables bound to states with `s', to name variables bound 
to goals with `g', and to name variables bound to search spaces with a suffix 
`-inf'. The first function, \texttt{disj2}, implements binary disjunction. 
\texttt{append-inf} is its helper, which compose two disjunctive search spaces. 
The following function, \texttt{conj2}, implements binary conjunction. It applys 
the \emph{first} goal to the input state, then applys the second goal to 
states in the resulting search space. The latter process is done with a helper 
function. \texttt{append-map-inf} applys its input goal to states in its input 
search spaces and compose the resulting search spaces. It reuse 
\texttt{append-inf} for search space composition. The following three 
definitions introduces syntactic sugars that miniKanren users are more familiar 
with. The first two definitions say disjunction and disjunction are 
right-associative. The next and last definition says \conde{} relates its 
clauses disjunctively, and goals in the same clause conjunctively.

@@

Balanced interleaving DFS (DFS$_\textrm{bi}$) has almost-fair \disj{} and unfair 
\conj{}. The implementation of DFS$_\textrm{bi}$ differs from DFS$_\textrm{i}$ 
in the \disj{} macro. We list the new \disj{} with its helpers in 
Fig.~\ref{balanced-disj}. The first helper function, \texttt{split}, takes a 
list of goals \texttt{ls} and a procedure \texttt{k}, partitions \texttt{ls} 
into two sub-lists of roughly equal length, and returns the application of 
\texttt{k} to the two sub-lists. \texttt{disj*} takes a non-empty list of goals 
\texttt{gs} and returns a goal. With the help of \texttt{split}, it essentially 
constructs a \emph{balanced} binary tree where leaves are elements of 
\texttt{gs} and nodes are \texttt{disj2}s, hence the name of this search 
strategy. In contrast, the \disj{} in DFS$_\textrm{i}$ constructs the binary 
tree with the same collection of leaf nodes but in a particularly unbalanced 
form.

@@

Fair DFS (\DFSf) has fair \disj{} and unfair \conj{}. The 
implementation of \DFSf{} differs from \DFSi{}'s in 
\texttt{disj2} (Fig.~\ref{fDFS}). \texttt{disj2} is changed to call a new and 
fair version of \texttt{append-inf}. \texttt{append-inf/fair} immediately calls 
its helper, \texttt{loop}, with the first argument, \texttt{s?}, set to 
\texttt{\#{}t}, which indicates that 
\texttt{s-inf} and \texttt{t-inf} haven't been swapped. The swapping happens at 
the third \texttt{cond} clause in the helper, where \texttt{s?} is updated 
accordingly. The first two \texttt{cond} clauses essentially copy the 
\texttt{car}s and stop recursion when one of the input spaces is obviously 
finite. The third clause, as we mentioned above, is just for swapping. When the 
fourth and last clause runs, we know that both \texttt{s-inf} and \texttt{t-inf} 
are ended with a thunk. In this case, a new thunk is constructed. The new thunk 
calls the driver recursively. Here changing the order of \texttt{t-inf} and 
\texttt{s-inf} won't hurt the fairness (though it will change the order of 
answers). We swap them back so that answers are produced in a more natural 
order.

@@

BFS has both fair \disj{} and fair \conj{}. Our implementation is based on 
DFS$_\textrm{f}$ (not DFS$_\textrm{i}$). To implement \BFS{} based on \DFSf{}, 
we need \texttt{append-map-inf/fair} in addition to \texttt{append-inf/fair}. 
The only difference between \texttt{append-map-inf/fair} and 
\texttt{append-map-inf} is that the former calls \texttt{append-inf/fair} 
instead of \texttt{append-inf}.
